|

### 2025-05-28 Reviews/round_1/rebuttal_point_by_point.md

> **The abstract sound great.**

Thanks for this nice comment.

> **The literature review provides a broad context for AI’s potential; however, it lacks a strong connection to recent studies that address AI’s evolving role in educational contexts, critical thinking, and argumentative writing. The paper could enhance its theoretical foundation by engaging more deeply with contemporary research on AI’s impact on students' critical thinking, argumentation and reasoning. There are much more recent studies by scholars such as Kazem Banihashem and that you can consult with.**

We thank the reviewer for this comment. We have now expanded the literature review into a dedicated section (p. xxx). We have elaborated on the evolving role of LLMs in education, focusing on the literature speficic to AI chatbots. The section discusses the main findings and points out to a few gaps in the literature that we aim to fill with our study. At the same time, given the novelty of the LLMs applications in education, the theoretical foundation for our work are necessarily limited.

@@

> **More information on the method section, the context, the design, and the participant characteristics are needed to give the readers a clear description of the study. Please elaborate on the background information of the characteristics of the participants that could influence the way learners engage with AI and critical thinking and argumentation. Acknowledging these limitations can guide future research in this area.**

We have revised the manuscript expanding the Methods section to improve the clarity of our experimental design and provide more details about the context. Specifically:

- The revised Methods section begins with a summary of our research design and setup before presenting the details in the relevant subsections.

- We now clearly justify the recruitment of the two schools (see footnote XXX), as in Reviewer 1's point discussed above.

- We also expanded the "Background Measures" sub-section into a new section "Student characteristics" tthat provides a detailed description of all the variables collected, including our questions on students' AI attitudes and experience, their academic habits and grades, and we have included a new sub-section discussing how we collected AI-student interaction metircs.

- Finally, in the Results section, we have revised Table 2  (p. XXX) to illustrate the distribution of participants characteristics across treatments, which provides a more precise information on the demographics and academic skills of our participants.

@@

> **The results show promising outcomes. However, the findings of this study are derived from self-reported data collected from participants through surveys. However, it is important to note that the literature suggests caution in heavily relying on such self-reported data, as perception does not always equate to actions (see: https://urldefense.com/v3/__https://doi.org/10.1080/02602938.2024.2345669__;!!DOxrgLBm!DU2xLNvI0Ke45t43rGQAI3BYa8bni0Io2JesBwINodtn8kzLHvspBE2my62mGZlAoKzFv6RFxgUv6FWyBQIbCg9ObQu4SA$ ). Therefore, this must be acknowledged as a limitation of the study, along with suggestions for future research to address this issue.**

Many thanks for this important point. We have discussed this limitation in the conclusions [TBA] "solely on students’ perceptions of their learning experiences may not accurately reflect their actual learning"

@@

- To strengthen the theoretical foundation of our study, we have expanded the literature review to more thoroughly justify the formulation of each research question.

- In line with scientific conventions, we have now explicitly stated all hypotheses in the Introduction, rather than introducing them only in the Results section. And the Results section has also been updated to link the analysis to each of the five research questions (see next point).

The revised research questions are as follows:

- RQ1: Do AI-generated explanations enhance students’ problem-solving performance in
school tasks?
- RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-
generated solutions?
- RQ3: Do student-AI interactions guided by the Socratic method promote better performance?
- RQ4: Do student-AI interactions guided by the Socratic method increase students' confidence in their answers? 
- RQ5: Do students perceive Socratic student-AI interactions more helpful than non-Socratic AI interactions?

@@

> **The results section first describes the group of subjects. For one experimental intervention (socratic AI tutor), it is reported whether the experimental group differed significantly from the control group. However, this information is missing for the other experimental intervention (AI explanation). In the remainder of the section, the analyses of the dependent variables are presented in a largely scientific manner. Notably, for the first experimental manipulation, a dependent variable has been listed in the methods section (propensity to update initial guess). However, no results are reported for this variable, and I do not understand what was done with this data or why it wasn't used. In the second experimental manipulation (socratic tutor), variables are reported that have not been described in the methods section. These new variables that measure the interaction with the AI tutor are only (and very briefly) motivated here. They should be motivated in the Introduction and included in the Methods section.**

Specifically:

- Regarding the missing information about the other experimental intervention, Table 2 now shows descriptives for both treatment groups. It also shows p-values from separate statistical tests of independence between variable and treatmnet assignment.

- Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. We report this analysis in footnote XXXX at p. xXXX.

- Regarding the missing information in the Methods section, we now report the student-AI interaction metrics -- logs and number of words -- in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p. xxxx).

@@

- In accordance to convention, we changed the sentence with "marginally significant (p = 0.15)" to "insignificant (p = 0.15)".

- We have also clarified the bootstrap resampling approach used to compute the difference in media accuracy between the treatmetn groups. Specifically:

> *Figure \@ref(fig:boot-accuracy) illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning. To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach. This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group. The 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups. The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).*

- In the Supporting Information section, we have clarified the information about the regression models [TBC]

### 2025-05-28 01-intro.Rmd
edits

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades [@roll2016evolution]. Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots]. Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of educational tasks: facilitating problem-solving [@urban2024chatgpt], language learning [@derakhshan2024chatgpt;@song2023enhancing], or evaluation of students' work [@henkel2024can]. However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges]. This study addresses two critical aspects of AI integration in education: the effectiveness of AI-generated explanations and the optimal modes of interaction between AI and students, specifically whether Socratic or alternative approaches yield better outcomes.

A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries. This study investigates how this ability influences students' interactions and helps them enhance their problem-solving skills. Specifically, we first examine the impact of AI explanations on students' performance in a numerical estimation task. We then test whether providing such AI explanations influence students' assessment of the accuracy of AI-generated predictions, compared to presenting the predictions alone. Finally, we test two approaches throught which students can interact with AI chatbots and receive explanations: one approach that encourages critical thinking through incremental guidance ("Socratic AI")  and another providing immediate solutions ("non-Socratic AI").

<!-- A key advancement over previous AI systems is that LLMs can generate step-by-step solutions to accompany their responses to students' queries. This study investigates how LLMs' ability to generate step-by-step explanations influences students' problem-solving skills and their critical evaluation of AI-generated responses. Specifically, we examine the impact of AI explanations on students' performance in a numerical estimation task and their assessment of the accuracy of AI-generated predictions. -->

Previous research has shown that explanations can foster children's causal learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions; @danovitch2021mind]. Similarly,  explanations generated by LLMs may guide students in their problem-solving, for example, by decomposing a numerical estimation task into more manageable steps, such as considering heuristics or approximations. However, LLMs can also generate inaccuracies and flawed reasoning [@saxton2019analysing; @kalyan2021how; @hendrycks2021measuring], misleading students if followed without scrutiny. Understanding this dual role of AI explanations as a valuable educational tool and a potential source of misinformation leads to our first research questions (RQs):

@@

Research in AI shows that providing explanations can increase users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability]. However, in educational settings, trust is more complex. Students may interpret well-written explanations as evidence that an answer is correct, even when it is not. Conversely, they may notice minor errors or inconsistencies in an explanation that reduce their trust, even if the overall solution is accurate. This motivates our second research question:

**RQ2: How do AI-generated explanations influence students’ perceived credibility of AI-generated solutions?**

The study also aims to explore how different modes of student-AI interaction more broadly influence students' performance and perceptions. Specifically, we contrast two fundamental approaches: the "Socratic” and “non-Socratic” methods. The Socratic approach, inspired by the Socratic teaching method [@lam2011socratic;@dalim2022promoting], engages students through an argumentative dialogue where the AI tutor asks thought-provoking questions to stimulate critical thinking and self-reflection [@wilberding2021socratic]. This method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning [@cleveland2015beyond]. In contrast, the non-Socratic approach directly answers students' queries without necessarily engaging them in a dialogue or exploration.

The Socratic approach has recently gained considerable attention in the context of AI tutoring. Notably, Khan Academy has developed its AI tutor, Khanmigo, based on these principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor]. Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation [@duelen2024socratic]. This debate raises the following research questions:

**RQ3: Do student-AI interactions guided by the Socratic method promote deeper critical thinking?**

@@

A key challenge when choosing between Socratic and non-Socratic AI is assessing the extent of "user satisfaction" among students, regardless of its academic benefits. If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI and seek (non-Socratic) tools that provide direct answers. Indeed, research on conversation length with LLM-powered Chatbots
has shown that people tend to have mixed reactions when chatbots are instructed to engage in longer conversations [@huang2024does]. Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools. This leads to our last research question:

@@

By focusing on the impact of AI explanations and the Socratic AI approach on students' critical thinking, we aim to contribute to a growing literature about integrating AI into education. Recent research on AI chatbots and LLMs indicates significant potential for improving student learning outcomes. A meta-analysis of 24 randomised studies highlights the positive impact of chatbots, especially on student productivity and learning engagement [@wu2024ai]. However, several studies are more cautious about the benefits, partly due to concerns that AI might encourage cheating, over-reliance, or superficial understanding, which are challenging to measure in short-term studies [@lee2024cheating]. Additional studies on LLMs' impact on academic integrity show that students and teachers struggle to differentiate between human and AI-generated text, raising risks of increased academic dishonesty and eroded trust [@yusuf2024generative; @eke2023chatgpt]. Furthermore, as LLMs continue to evolve, students and educators may lack the skills to leverage them effectively, which could result in missed educational opportunities [@kaplan2023generative].

<!-- The ethical and societal implications of LLMs in education are also complex. Issues such as bias, privacy, surveillance, and autonomy persist, especially in K-12 contexts, where the impact of algorithmic decisions on young learners' agency and privacy can be profound [@williams2024ethical]. LLMs introduce unique challenges, notably in alignment, where the system's goals may not align fully with educational objectives, despite companies' efforts to mitigate such risks through fine-tuning and reinforcement learning [@gabriel2020artificial]. Concerns also extend to students' digital well-being, as AI in education could inadvertently contribute to unhealthy digital habits and reliance on AI assistance over critical thinking [@weidinger2022taxonomy]. Given these multidimensional impacts, further research is needed to optimise AI's role in education, balancing short-term learning gains with long-term skill development and ethical considerations. -->

<!-- Contributions -->

Research in the field of human-AI combinations does not indicate consensus in terms of its effectiveness. For example, using a meta-analysis of studies published between 2020 and 2023, @vaccaro2024combinations finds that, on average, human-AI combinations performed significantly worse than the best of humans or AI alone. On the other hand, @li2024explanatory uses a sample of 200 Chinese students to show significant correlations between human–computer interactions, perceived usefulness and students' skills. Understanding these associations can help institutions design better AI courses and more effective integration of AI in education.

@@

Results of our investigation also contribute to the current debate on the pedagogical principles for designing AI-driven systems and educational opportunities, the so-called _design for learning_ [@gavsevic2023empowering]. It underscores the importance of using a multi-disciplinary approach that combines traditional pedagogical insights with principles from human-computer interactions. Building on recent efforts to address critical challenges in integrating AI into education ---like providing feedback, grading, and making learning recommendations [@dai2023can; @yan2024practical; @joksimovic2023opportunities]--- this study offers insights on how k-12 students engage with alternative forms of AI interaction. In particular, it emphasizes the challenges of embedding classical pedagogical principles, such as the Socratic Method, and the value of providing AI explanations for students compared to just solutions.

<!-- One concern is that students will rely on AI results instead of learning from it, resulting in an ineffective AI human combination. In a related experimental study, @darvishi2024impact (n = 1625)  -->

>>>>

Integrating artificial intelligence (AI) into educational settings has been debated for over three decades [@roll2016evolution]. Recent advances in generative AI and large language model (LLM) chatbots have opened even more possibilities for improvements in educational practices [@yan2024promises;@kasneci2023chatgpt;@tlili2023if;@debets2025chatbots]. Although LLMs mimic human intelligence, their applications extend beyond imitation and include a range of academic tasks: facilitating problem-solving [@urban2024chatgpt], providing personalised assistance [@darvishi2024impact], language learning [@derakhshan2024chatgpt;@song2023enhancing], evaluation of students' work [@henkel2024can]. However, the impact of these applications on students' learning remains controversial, with proponents highlighting potential benefits and critics cautioning about risks and drawbacks [@farrokhnia2024swot;@weidinger2022taxonomy;@walczak2023challenges].

A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries [@wei2022chain]. Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions;@danovitch2021mind], there is limited empirical research examining how such AI-generated explanations impact students' learning processes. Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills. To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) to address our first research question (RQ):

@@

The distinction between AI-generated *explanations* and *solutions* further allows us to evaluate whether providing explanations influences students' trust and judgment of AI-generated content. Indeed, LLMs can generate inaccurate solutions and flawed reasoning [@saxton2019analysing;@kalyan2021how;@hendrycks2021measuring]. However, there is evidence that providing explanations increases users' trust and acceptance of AI systems, especially when the reasoning behind the output is clear [@rai2020explainable;@ferrario2022explainability]. In an education setting, this may affect learning outcomes in a complex manner. Students may interpret well-written explanations as evidence that an answer is correct, even when it is not. Conversely, they may notice minor errors or inconsistencies in an explanation that reduces their trust, even if the overall solution is accurate. This situation motivates our second research question:

**RQ2: How do AI-generated explanations influence students' perceived credibility of AI-generated solutions?**

Another critical technical progress of contemporary LLMs over previous AI systems is their ability to follow specific instructions when engaged in conversations.  Building on this capability, another goal of this study is to examine the impact of two fundamental modes of interaction on students' learning outcomes: one that fosters critical thinking through incremental guidance ("Socratic AI") and another that offers immediate solutions ("non-Socratic AI").

Although classical pedagogical methods like the Socratic Method have long been valued for promoting critical thinking and more profound understanding [@lam2011socratic;@dalim2022promoting;@wilberding2021socratic;@cleveland2015beyond], there is a notable gap in the AI education literature regarding how to integrate these methods into AI-driven solutions. Existing AI tools, like ChatGPT, often focus on providing direct solutions or feedback, potentially limiting opportunities for active student engagement and reflection. Our study addresses this gap by conducting a parallel experimental intervention comparing Socratic-style AI interactions to more conventional, solution-focused AI interactions in a K-12 setting. This comparison contributes novel evidence on how different AI interaction modes can influence student learning processes and outcomes, informing the design of more pedagogically sound AI educational systems.

More specifically, building on previous literature suggesting that the Socratic method encourages students to think critically and arrive at their answers rather than relying on the AI-generated solution, thus helping students gain more confidence in their reasoning [@cleveland2015beyond], we test the following research questions:

**RQ3: Do student-AI interactions guided by the Socratic method promote better performance?**

@@

In addition to educational outcomes, exploring the trade-offs between Socratic and non-Socratic AI raises an important question about student *satisfaction*. If students find interacting with Socratic AI chatbots unengaging or frustrating, such perceptions might drive students away from Socratic AI to seek other (non-Socratic) tools. Indeed, research on conversation length with LLM-powered chatbots has shown that people tend to have mixed reactions when chatbots are instructed to engage in more extended conversations [@huang2024does]. Therefore, aligning educational benefits with user satisfaction is crucial for ensuring the long-term adoption of AI learning tools. This consideration leads to our last research question:

@@

<!-- In this study, we use a randomised experiment to investigate these questions. For  by collecting detailed information about students' conversation logs and length with LLM-based chatbot, which we combine with self-reported data -->

Addressing these questions aims to contribute to a growing debate on using pedagogical principles, like a Socratic approach, to integrate AI in education, which has recently gained considerable attention. For instance, Khan Academy has developed its AI tutor, Khanmigo, based on Socratic principles, aiming to foster critical thinking and engagement in learners.^[https://docsbot.ai/prompts/education/khanmigo-lite-socratic-tutor] Several commentators have argued that the Socratic method is particularly effective for designing AI tutors for children, as it enhances critical thinking and discourages students from copying AI-generated responses without scrutiny [@lara2020artificial].^[See also https://research.gatech.edu/ai-oral-assessment-tool-uses-socratic-method-test-students-knowledge] Beyond education, Socratic AI systems are also being developed to foster critical thinking among the general public, particularly in efforts to combat disinformation [@duelen2024socratic].

## Literature Review

A growing body of research highlights the tremendous potential of LLMs to improve student learning outcomes. A recent scoping review identifies 53 distinct use cases for LLMs in education [@yan2024practical], ranging from automatic grading and personalised feedback to teaching support and content generation, reflecting the high versatility and scalability of LLMs. Within this expanding literature, this study centers on *AI chatbots*, conversational agents powered by LLMs, and their educational applications, reflecting their increasing prominence and potential impact in this field.

Early investigations into the use of AI chatbots, such as ChatGPT, in educational settings have identified both opportunities and challenges [@farrokhnia2024swot;@noroozi2024generative]. On one hand, ChatGPT can serve as a personalised tutor, assist in drafting assignments, or generate creative prompts for classroom activities. On the other hand, concerns have been raised around academic integrity, overreliance on AI-generated content, and the need for critical thinking skills when interpreting chatbot outputs.

A recent review of over 70 studies offers a broad look at how chatbots are used in different areas of education [@debets2025chatbots]. It groups chatbot uses by roles like tutoring, giving feedback, and helping with administrative tasks, and examines how they are designed and used. The review finds that AI chatbots generally meet their goals. For example, a quasi-experimental study with 320 middle schoolers found that an AI teaching chatbot significantly improved math knowledge [@xing2025development]. Similarly, a meta-analysis of 24 RCTs shows that AI chatbots significantly enhance student outcomes, with stronger effects in higher education [@wu2024ai]. However, most studies are short-term, may suffer from publication bias, and often lack a strong theoretical foundation.

Despite promising results, key limitations remain. For instance, @er2024assessing found that students in a programming class performed better with instructor feedback, which was seen as more useful and fair than AI-generated feedback, highlighting the current gap between AI and human instruction. Simiarly, evidence of performance boosts from non-education settings is mixed. A meta-analysis by @vaccaro2024combinations found that human-AI teams often performed worse than either humans or AI alone, suggesting a more nuanced relationship. In education, additional concerns are that intagrating AI might encourage cheating, over-reliance, or superficial understanding [@farrokhnia2024swot;@yusuf2024generative;@eke2023chatgpt;@lee2024cheating], the risk of which is hard to evaluate without longitudinal studies.

The effectiveness of AI in education often depends on students' attitudes and perceived usefulness of AI tools, understanding which can guide curriculum design. For instance, @li2024explanatory found that K–12 students’ views on AI tools were linked to factors like readiness, confidence, and anxiety. Similar concerns have been noted from teachers' perspectives as well [@kaplan2023generative]. Our study builds on this by exploring how interaction styles, such as Socratic vs. non-Socratic, shape students' perceptions of AI helpfulness.

@@

Most existing research focuses on general AI chatbots, like ChatGPT, which are not specifically designed for students and often lack clear pedagogical foundations. Our study highlights the importance of purposeful design and pedagogical goals, contributing to the growing field of AI chatbot design for education, or design for learning [@gavsevic2023empowering]. This research calls for a multidisciplinary approach combining pedagogy and human-computer interaction. Building on recent efforts [@dai2023can; @yan2024practical; @joksimovic2023opportunities], we examine how K–12 students interact with different AI chatbot styles. By doing so, thee study underscores both the potential and challenges of applying classical pedagical methods, such as the Socratic approach, emphasizing the need to provide explanations rather than just answers.
